---
title: "Custom Pathways for Non-Model Organisms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Pathways for Non-Model Organisms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

easyGSEA supports custom pathway annotations for non-model organisms that don't have pre-built annotation databases (org.*.eg.db packages).

This vignette shows you how to create and use custom pathway files.

## When to Use Custom Pathways

**Use custom pathways when:**
- Working with non-model organisms (e.g., *Bradysia*, emerging model systems)
- Standard organism databases don't exist
- You have species-specific pathway annotations
- You want to use custom gene sets (e.g., from literature)

**Don't need custom pathways if:**
- Your organism is: human, mouse, rat, fly, zebrafish, worm, yeast
- Use `organism = "auto"` instead (easier!)

## File Formats

easyGSEA accepts two formats:

### CSV Format (Recommended)

Simple two-column format:
```
pathway,gene_id
DNA_repair,gene_001
DNA_repair,gene_002
DNA_repair,gene_005
Cell_cycle,gene_003
Cell_cycle,gene_004
Cell_cycle,gene_007
Translation,gene_006
Translation,gene_008
```

**Requirements:**
- Header row with columns: `pathway` and `gene_id`
- One gene per row
- Genes can appear in multiple pathways
- Pathway names: use underscores (not spaces)
- Gene IDs: match your DESeq2 rownames EXACTLY

### GMT Format (Advanced)

Standard gene set format:
```
DNA_repair	description	gene_001	gene_002	gene_005
Cell_cycle	description	gene_003	gene_004	gene_007
Translation	description	gene_006	gene_008
```

**Format:**
- Tab-separated
- Column 1: Pathway name
- Column 2: Description (can be "NA")
- Column 3+: Gene IDs

## Creating Your Pathway File

### Step 1: Identify Your Gene IDs

Your gene IDs must match DESeq2 rownames:
```{r check-ids}
library(DESeq2)

# Check your gene IDs
head(rownames(dds))
# Example output: "Brad_gene_001" "Brad_gene_002" ...
```

### Step 2: Map Genes to Pathways

**Option A: Use Orthology**

Map your genes to a model organism, inherit pathways:
```{r orthology}
# 1. Find orthologs (e.g., using BLAST or OrthoFinder)
# 2. Get pathways for model organism orthologs
# 3. Transfer pathways to your genes

# Example result:
# Your_gene_001 → Drosophila FBgn123 → GO:0006281 (DNA repair)
```

**Option B: Literature-Based**

Manually curate from publications:
```{r manual}
# Create data frame
pathways <- data.frame(
  pathway = c("DNA_repair", "DNA_repair", "Cell_cycle"),
  gene_id = c("Brad_gene_001", "Brad_gene_002", "Brad_gene_003")
)

# Save as CSV
write.csv(pathways, "bradysia_pathways.csv", row.names = FALSE)
```

**Option C: Functional Annotation**

Use annotation tools:
```{r annotation}
# Tools to consider:
# - InterProScan (protein domains)
# - eggNOG-mapper (orthology + GO terms)
# - BLAST2GO (BLAST + GO mapping)
# - Trinotate (for transcriptomes)

# These generate GO terms you can convert to pathway file
```

## Example Workflow: Bradysia coprophila

Real example from non-model organism research:

### Step 1: Prepare Gene List
```{r bradysia-prep}
library(easyGSEA)
library(DESeq2)

# Load your DESeq2 object
# Genes are named: Brad_01, Brad_02, etc.
head(rownames(dds))
```

### Step 2: Create Pathway File

Using orthology to Drosophila:
```{r bradysia-orthology}
# Find Drosophila orthologs (external tool: OrthoFinder, BLAST)
# Map Drosophila genes to GO terms
# Transfer GO terms to Bradysia genes

# Example pathway file:
bradysia_go <- data.frame(
  pathway = c(
    "GO:0007049", "GO:0007049", "GO:0007049",  # Cell cycle
    "GO:0006281", "GO:0006281",                 # DNA repair
    "GO:0006412", "GO:0006412", "GO:0006412"   # Translation
  ),
  gene_id = c(
    "Brad_001", "Brad_045", "Brad_089",
    "Brad_012", "Brad_067",
    "Brad_023", "Brad_098", "Brad_134"
  )
)

# Save
write.csv(bradysia_go, "bradysia_pathways.csv", row.names = FALSE)
```

### Step 3: Run GSEA
```{r bradysia-gsea}
# Run with custom pathways
results <- gsea_from_deseq(
  dds = dds,
  contrast = c("stage", "germline", "soma"),
  custom_pathways = "bradysia_pathways.csv"
)

# Results are same format as standard GSEA
nrow(results)
head(results@result[, c("Description", "NES", "p.adjust")])
```

### Step 4: Visualize
```{r bradysia-plot}
# Same plotting functions work
plot_gsea(results)

# Compare developmental stages
stage1 <- gsea_from_deseq(dds, 
                         contrast = c("stage", "stage1", "control"),
                         custom_pathways = "bradysia_pathways.csv")
                         
stage2 <- gsea_from_deseq(dds,
                         contrast = c("stage", "stage2", "control"),
                         custom_pathways = "bradysia_pathways.csv")

comparison <- compare_gsea(stage1, stage2, 
                          label1 = "Stage 1", label2 = "Stage 2")
plot_comparison(comparison)
```

## Validation Tips

### Check Your File
```{r validate}
# Read your file
pathways <- read.csv("my_pathways.csv")

# Check format
colnames(pathways)  # Should be: "pathway" "gene_id"

# Check gene IDs match DESeq2
gene_overlap <- intersect(pathways$gene_id, rownames(dds))
coverage <- length(gene_overlap) / nrow(dds) * 100

cat("Gene coverage:", round(coverage, 1), "%\n")
# Aim for >70% coverage
```

### Pathway Size Guidelines
```{r pathway-size}
# Count genes per pathway
pathway_sizes <- table(pathways$pathway)

# Check distribution
summary(as.numeric(pathway_sizes))

# Recommendations:
# - Minimum: 10 genes per pathway (smaller = unreliable)
# - Maximum: 500 genes per pathway (larger = too broad)
# - Ideal: 20-200 genes per pathway
```

### Common Issues

**Issue 1: No pathways found**
```{r issue1}
# Check gene ID matching
setdiff(pathways$gene_id[1:10], rownames(dds))
# Should be empty. If not, IDs don't match!
```

**Issue 2: Very few pathways**

- Need more pathway annotations
- Try relaxing padj_cutoff: `padj_cutoff = 0.1`
- Check if knockdown is strong enough

**Issue 3: All pathways enriched**

- Pathway file too small/broad
- Integration artifacts dominating
- Check with replicate comparisons

## Tips for Different Organisms

### For Transcriptome Assemblies
```{r transcriptome}
# Use Trinity gene IDs directly
# GENE_001, GENE_002, etc.

# Annotate with Trinotate
# Extract GO terms
# Convert to pathway file
```

### For Genome Annotations
```{r genome}
# Use official gene IDs
# e.g., LOC genes, maker annotations

# Use eggNOG-mapper for GO terms
# Download results
# Parse into pathway file
```

### For Cell Lines / Variants
```{r cellline}
# Use human/mouse gene IDs
# But create custom pathways for:
# - Cell-type specific processes
# - Disease-relevant pathways
# - Literature-curated gene sets
```

## Resources

**Orthology Tools:**
- OrthoFinder: https://github.com/davidemms/OrthoFinder
- eggNOG-mapper: http://eggnog-mapper.embl.de/
- OrthoDB: https://www.orthodb.org/

**Functional Annotation:**
- InterProScan: https://www.ebi.ac.uk/interpro/
- Trinotate: https://github.com/Trinotate/Trinotate
- BLAST2GO: https://www.blast2go.com/

**Gene Ontology:**
- GO database: http://geneontology.org/
- QuickGO: https://www.ebi.ac.uk/QuickGO/

## Example Files

See package GitHub for example pathway files:
- `inst/extdata/bradysia_pathways.csv`
- `inst/extdata/example_pathways.gmt`

## Next Steps

- See `vignette("quickstart")` for basic usage
- See `vignette("comparison")` for comparing conditions
- Check `?gsea_from_deseq` for all parameters

## Citation

If you use easyGSEA with custom pathways, please cite:

- **easyGSEA**: Samuel GH (2026). easyGSEA: Simplified Gene Set Enrichment Analysis. R package version 0.1.0.

- **clusterProfiler**: Yu G, Wang LG, Han Y, He QY (2012). "clusterProfiler: an R package for comparing biological themes among gene clusters." OMICS, 16(5), 284-287.
